Rational Monte Carlo Sampling in Trees

### 2012-01-11 paper/rmcts/paper.tex
paper in progress

\title{Doing Better Than UCT: \\ Rational Monte Carlo Sampling in Trees}
\author {David Tolpin, Solomon Eyal Shimony \\
Department of Computer Science, \\
Ben-Gurion University of the Negev, Beer Sheva, Israel \\
\{tolpin,shimony\}@cs.bgu.ac.il}

@@

The ultimate goal of this ongoing work is ``optimal'' sampling
in the meta-reasoning sense. Nevertheless, this task is daunting for
the following reasons:
\begin{itemize}
\item Defining the cost of a sample is not easy, and even if we simply use time-cost
as an approximation, we get an intractable meta-reasoning problem.
\item Applying the standard myopic and subtree independence assumptions, we run into
serious problems. Even in the standard selection problem \cite{TolpinShimony.blinkered}, we get a non-concave
utility function and premature stopping of the algorithm. This is due to the fact
that the value of information of a single measurement (analogous
to a sample in MCTS) is frequently less than its
time-cost, even though this is not true for multiple measurements.
When applying the selection problem to MCTS, the situation is exacerbated. 
The utility of an action is usually bounded, and thus in many cases a single sample
may be insufficient to change the current best action, {\em regardless} of its
outcome. As a result, we frequently
get a {\em zero} ``myopic'' value of information for a single sample.
\end{itemize}

As the ultimate task is extremely difficult
to achieve, and even harder to analyze, we begin with
simple schemes more amenable to analysis. We then
develop a crude value of information (VOI)
based scheme. Although the latter is a somewhat ad-hoc attempt to
estimate value of information of a sample, it already appears (empirically) to do
better than UCT as well as better than all the simpler schemes we propose.

\section{Main Results}

\subsection{Doing better than UCB}

\begin{enumerate}
\item The $\varepsilon$-greedy allocation scheme has exponential convergence
rate of the simple regret $\IE r$ (see Appendix~\ref{app:derivations-finite-eps}):
\begin{equation}
  \IE r_\varepsilon\le\sum_{j=1}^K\Delta_j\left(\frac {n\varepsilon} K + \frac {4\sqrt e}
{\Delta_j^2}\right)e^{-\Delta_j^2n\varepsilon/8K}
\end{equation}
\item The UCB allocation scheme exhibits polynomial convergence
rate (see Appendix~\ref{app:derivations-finite-ucb}):
\begin{equation}
\IE r_{ucb} \le 2\sum_{j=1}^K \Delta_jn^{\frac {-\alpha \Delta_j^2} 8}
\end{equation}
\item A tighter upper regret bound for the $\varepsilon$-greedy allocation scheme depends on
$\varepsilon$. The lowest bound is
achieved for $\varepsilon=\frac 1
2$, and for a large number of arms the bound for the $\frac 1 2$-greedy scheme approaches the square of the
bound for uniform sampling (see
Appendix~\ref{app:derivations-asym-eps}).

@@

\subsection{Doing better than UCT}

@@

\section{Empirical Evaluation}

The results were empirically verified on Multi-armed Bandit instances
(Section~\ref{seq:emp-mab}), on search trees
(Section~\ref{seq:emp-mcts}), and on the sailing domain
(Section~\ref{seq:emp-sailing}), as defined in \cite{Kocsis.uct}. The
experiments confirmed the hypotheses. The algorithms used in the
experiments are:
\begin{description}
\item[RND:] uniform \underline{r}a\underline{nd}om samping;
\item[UCT:] \underline{U}pper \underline{C}onfidence Bounds applied to
  \underline{T}rees \cite{Kocsis.uct};
\item[GCT:] U\underline{CT} with the first step
  replaced with $\frac 1 2$-\underline{g}reedy sampling;
\item[QCT:] U\underline{CT} with the first step
  replaced with U\underline{Q}B.
\end{description}

\subsection{Simple regret in multi-armed bandits}
\label{seq:emp-mab}

Figure~\ref{fig:mab-simple-regret} presents a comparison of simple
regret minimization algorithms for Multi-armed
bandits. Figure~\ref{fig:mab-simple-regret}.a shows the search tree
corresponding to a problem instance. Each arm returns a random reward
drawn from a Bernoulli distribution. The search selects an arm
and compares the expected reward, unknown to the algorithm during the
sampling, to the expected reward of the best arm.

@@

For smaller numbers of samples, $\frac 1 2$-greedy achieves the best
performance; for large number of samples, QCT outperforms GCT. QCT is
better than UCT everywhere except for very small numbers of samples. A
combination of GCT and QCT dominates UCT over the whole range.

@@

\subsection{Monte Carlo tree search}
\label{seq:emp-mcts}

@@

\subsection{The sailing domain}
\label{seq:emp-sailing}

@@

\section{Summary and Future Work}

@@

VOI of a rollout can be computed when the sample distribution of an
action is known up to the parameters, such as the normal distribution
with an unknown mean and/or variance. Alternatively, the value of
information can be estimated from the set of samples, and the need to
assume a particular shape of the distribution can be lifted. In one
realization of the latter approach the VOI of performing an action is
estimated as \emph{the value of perfect information learned from the outcome
of earlier samples of the action divided by the number of
samples}. Early experiments with this approach showed the best
performance for sets of Bernoulli arms, and even a more prominent
improvement for mixed arms (e.g. Bernoulli, triangularly distributed,
and fixed arms) (Figure~\ref{fig:flat-trilevel-vct}):
\textbf{VCT}, the algorithm which selects an action with the maximum VOI estimate on
the first step exhibits lower average regret than other algorithms.
Sample-based VOI estimation requires keeping more information than an
algorithm based only on the sample
mean, but the overhead is small and justified by the improvement.

>>>>

\title{Rational Monte Carlo Sampling in Trees}
\author {David Tolpin, Solomon Eyal Shimony \\
Department of Computer Science, \\
Ben-Gurion University of the Negev, Beer Sheva, Israel \\
\{tolpin,shimony\}@cs.bgu.ac.il}

@@

We begin with introduction of several sampling schemes with
better bounds for the simple regret on sets (Section~\ref{sec:sampling-on-sets}). We then
extend the resutls to sampling in trees by combining the proposed sampling
schemes on the first step of a rollout with UCT during the rest of the
rollout (Section~\ref{sec:sampling-in-trees}). Finally, we empirically
evaluate the performance of the proposed sampling schemes on sets of
Bernoulli arms (Section~\ref{sec:emp-mab}), in randomly generated
trees (Section~\ref{sec:emp-mcts}), and on the sailing domain
(Section~\ref{sec:emp-sailing}). Finally, we describe future research
directions and provide preliminary results for a sampling scheme
explicitly based on the value of information of a rollout (Section~\ref{sec:summary}).

\section{Related Work}
\label{sec:related-work}

Efficient algorithms for Multi-Armed Bandits based on
distribution-independent bounds, in particular UCB1, are introduced in
\cite{Auer.ucb}. The UCT algorithm, an extension of UCB1 to
Monte-Carlo Tree Search is described in \cite{Kocsis.uct}. Pure
exploration in Multi-armed bandits is explored in
\cite{Bubeck.pure}. On the one hand, the paper proves certain upper
and lower bounds for UCB1 and uniform sampling, showing that an upper
bound on the simple regret is exponential in the number of samples for
uniform sampling, while only polynomial for UCB1. On the other hand,
empirical performance of UCB1 appears to be better than of uniform
sampling.

The principles of bounded rationality appear in
\cite{Horvitz.reasoningabout}. \cite{Russell.right} provided a formal
description of rational metareasoning and case studies of applications
in several problem domains. One obstacle to straightforward
application of the principles of rational metareasoning to Monte-Carlo
sampling is the metagreedy assumption, according to which samples must
be selected as though at most one sample can be taken before an action
is chosen. In Monte-Carlo sampling, the value of information of any
single sample in a given search state is often zero, so a different
approximating assumption must be used instead.

\section{Main Results}
\label{sec:results}

The ultimate goal is ``optimal'' sampling in the meta-reasoning
sense. Nevertheless, this task is daunting for the following reasons:
\begin{itemize}
\item Defining the cost of a sample is not easy, and even if we simply
  use time-cost as an approximation, we get an intractable
  meta-reasoning problem.
\item Applying the standard myopic and subtree independence
  assumptions, we run into serious problems. Even in the standard
  selection problem \cite{TolpinShimony.blinkered}, we get a
  non-concave utility function and premature stopping of the
  algorithm. This is due to the fact that the value of information of
  a single measurement (analogous to a sample in MCTS) is frequently
  less than its time-cost, even though this is not true for multiple
  measurements.  When applying the selection problem to MCTS, the
  situation is exacerbated.  The utility of an action is usually
  bounded, and thus in many cases a single sample may be insufficient
  to change the current best action, {\em regardless} of its
  outcome. As a result, we frequently get a {\em zero} ``myopic''
  value of information for a single sample.
\end{itemize}

As the ultimate task is extremely difficult
to achieve, and even harder to analyze, we introduce
simple schemes more amenable to analysis, and compare
them to UCB (on sets) and UCT (in trees).

\subsection{Sampling on Sets}
\label{sec:sampling-on-sets}

\begin{enumerate}
\item The $\varepsilon$-greedy allocation scheme has exponential convergencerate of the simple regret $\IE r$ (see Appendix~\ref{app:derivations-finite-eps}):
\begin{equation}
  \IE r_\varepsilon\le\sum_{j=1}^K\Delta_j\left(\frac {n\varepsilon} K + \frac {4\sqrt e}
{\Delta_j^2}\right)e^{-\Delta_j^2n\varepsilon/8K}
\end{equation}
\item The UCB allocation scheme exhibits polynomial convergence
rate (see Appendix~\ref{app:derivations-finite-ucb}):
\begin{equation}
\IE r_{ucb} \le 2\sum_{j=1}^K \Delta_jn^{\frac {-\alpha \Delta_j^2} 8}
\end{equation}
\item An asymptotic upper bound on the simple regret for the
$\varepsilon$-greedy allocation scheme non-monotonically depends on
$\varepsilon$. The lowest bound is achieved for $\varepsilon=\frac 1
2$, and for a large number of arms the bound for the $\frac 1
2$-greedy scheme approaches the square of the bound for uniform
sampling (see Appendix~\ref{app:derivations-asym-eps}).

@@

\subsection{Sampling in Trees}
\label{sec:sampling-in-trees}

@@

\section{Empirical Evaluation}
\label{sec:emp}

The results were empirically verified on Multi-armed Bandit instances
(Section~\ref{sec:emp-mab}), on search trees
(Section~\ref{sec:emp-mcts}), and on the sailing domain
(Section~\ref{sec:emp-sailing}), as defined in \cite{Kocsis.uct}. The
experiments confirmed the hypotheses. The algorithms used in the
experiments are:
\begin{description}
\item[RND:] uniform \underline{r}a\underline{nd}om samping;
\item[UCT:] \underline{U}pper \underline{C}onfidence Bounds applied to
  \underline{T}rees \cite{Kocsis.uct};
\item[GCT:] U\underline{CT} with the first step
  replaced with $\frac 1 2$-\underline{g}reedy sampling;
\item[QCT:] U\underline{CT} with the first step
  replaced with U\underline{Q}B.
\end{description}

\subsection{Simple regret in multi-armed bandits}
\label{sec:emp-mab}

Figure~\ref{fig:mab-simple-regret} presents a comparison of MCTS sampling
schemes on Multi-armed bandits. Figure~\ref{fig:mab-simple-regret}.a shows the search tree
corresponding to a problem instance. Each arm returns a random reward
drawn from a Bernoulli distribution. The search selects an arm
and compares the expected reward, unknown to the algorithm during the
sampling, to the expected reward of the best arm. For such problem
instances, the MCTS sampling schemes coincide with the selection
algorithms employed at the first step of a rollout.

@@

For smaller numbers of samples, GCT achieves the best
performance; for large number of samples, QCT outperforms GCT. QCT is
better than UCT everywhere except for very small numbers of samples. A
combination of GCT and QCT dominates UCT over the whole range.

@@

\subsection{Monte Carlo tree search}
\label{sec:emp-mcts}

@@

\subsection{The sailing domain}
\label{sec:emp-sailing}

@@

\section{Summary and Future Work}
\label{sec:summary}

@@

VOI of a rollout can be computed when the sample distribution of an
action is known up to the parameters, such as the normal distribution
with an unknown mean and/or variance. Alternatively, the value of
information can be estimated from the set of samples, and the need to
assume a particular shape of the distribution can be lifted. In one
realization of the latter approach the VOI of performing an action is
estimated as \emph{the value of perfect information learned from the
  outcome of earlier samples of the action divided by the number of
  samples}. Early experiments with this approach showed the best
performance for sets of Bernoulli arms, and even a more prominent
improvement for mixed arms (e.g. Bernoulli, triangularly distributed,
and fixed arms) (Figure~\ref{fig:flat-trilevel-vct} shows results for
different algorithms for 32 randomly generated arms): \textbf{VCT},
the algorithm which selects a rollout with the maximum VOI estimate
exhibits lower average regret than other algorithms. Sample-based VOI
estimation requires keeping more information than an algorithm based
only on the sample mean, but the overhead is small and justified by
the improvement.
